Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-04 paper/main.tex
Updates paper

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on their domain knowledge. This preference arises due to several
practical issues with automated algorithms, such as they make obvious errors
and out Markov Equivalence Classes (MECs). To address these challenges, we
propose an iterative method to assist researchers in manually constructing or
improving an existing DAG by combining implied Conditional Independence (CI)
testing with expert knowledge. Our proposed method utilizes a novel measure of
partial association for mixed data to rank violations of implied CIs in a given
DAG, enabling users to prioritize modifications while utilizing their domain
knowledge to decide the orientation of new edges. Empirical results show that
this manual model construction approach achieves performance comparable to
automated algorithms even if the user is able to decide the edge orientation
correctly in one out of three cases. Additionally, we demonstrate that, in the
absence of an expert, a Large Language Model (LLM) can be potentially determine
edge orientations. We provide the implementation of our approach in both a web 
tool at: <redacted for review> and a Python package <redacted for review>.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are asymptotically
		consistent, their behavior on finite samples is not well
		understood. Additionally, their output can vary significantly
		depending on the choice of algorithm and hyperparameters,
		making it difficult to assess reliability. The absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. 
		% As a result,
		% even when using these algorithms are used, researchers often
		% rely on post hoc manual modifications to the model based on
		% model testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. 
		% As
		% a result, researchers have to manually orient these edges based
		% on their expert knowledge before using them for analysis.
\end{enumerate}

\todo[inline]{The following three paragraphs need work.}

Fig.~\ref{fig:intro} highlights some of these issues. To deal with these issues
methods for integrating expert knowledge have been developed. The two main
methods to integrate expert knowledge in the model construction process is to:
1. Provide all the expert knowledge to the causal discovery algorithm for
example in the form of required and forbidden edges \citep{Meek1995}, or
temporal information \citep{Bang2023} and let the algorithm use it to construct
the model. 2. Utilize conditional independence (CI) testing to identify issues in
the model, and manually fix them based on expert knowledge \citep{Ankan2021}.
We focus on this second approach in this paper.

@@

@@

\section{Web Tool}
\label{sec:web}
Based on the procedure outlined in the previous section, we built a web-based
tool that allows us to interactively create/modify their model. The user
starts by uploading their dataset and optionally the initial DAG. If no initial
DAG is provided an empty graph is initialized. The tool also computes the
Shipley's C value at any point. The web-tool would automatically show all
unexplained associations in the DAG by adding red edges between them. The edge
weight of the edge represents how high this strength is. Users can also set the
threshold for p-value and association to control how many edges are shown at a
time. Based on these shown potential edges, the user can select the edge that
they want to add and decide the direction for it. If the tool detects an
incorrect edge in the graph, it is shown in black. Once the user is happy with
the graph they have they can export it into different formats to use it for
further computation.

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on their domain knowledge. This preference arises due to several
practical issues with automated algorithms, such as they make obvious errors
and output Markov Equivalence Classes (MECs). To address these challenges, we
propose an iterative method to assist researchers in manually constructing or
improving an existing DAG by combining implied Conditional Independence (CI)
testing with expert knowledge. Our proposed method utilizes a novel measure of
partial association for mixed data to rank violations of implied CIs in a given
DAG, enabling users to prioritize modifications while utilizing their domain
knowledge to decide the orientation of new edges. Empirical results show that
this manual model construction approach achieves performance comparable to
automated algorithms if the user is able correctly orientation edges in one out
of three cases. Additionally, we demonstrate that, in the absence of an expert,
a Large Language Model (LLM) can be potentially determine edge orientations. We
provide the implementation of our approach in both a web tool at: <redacted for
review> and a Python package <redacted for review>.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are asymptotically
		consistent, their behavior on finite samples is not well
		understood. Additionally, their output can vary significantly
		depending on the choice of algorithm and hyperparameters,
		making it difficult to assess reliability. The absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. 
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. 
\end{enumerate}

Fig.~\ref{fig:intro} highlights some of these issues. When constructing models
manually it is important to test whether the constructed DAG is representative
of the dataset or not. Based on this testing and domain knowledge further
modifications can be made to the DAG. One of the approaches to test and modify
DAGs is by testing whether the implied CI tests of the DAG are true in the data
or not \citep{Ankan2021}. If some of the tests are violated in the data,
researchers can use their domain knowledge to make modifications to the DAG.

This CI testing

% To deal with these issues
% methods for integrating expert knowledge have been developed. The two main
% methods to integrate expert knowledge in the model construction process is to:
% 1. Provide all the expert knowledge to the causal discovery algorithm for
% example in the form of required and forbidden edges \citep{Meek1995}, or
% temporal information \citep{Bang2023} and let the algorithm use it to construct
% the model. 2. Utilize conditional independence (CI) testing to identify issues in
% the model, and manually fix them based on expert knowledge \citep{Ankan2021}.
% We focus on this second approach in this paper.

@@

\todo[inline]{Remove or move equations to appendix if no space}

@@

\section{Web Tool}
\label{sec:web}

To allow users to apply this method to their own dataset, we also built an
interactive web browser based tool to allow them to interactively construct
models. The user starts by uploading their dataset which initializes an empty
DAG on the variables in the dataset. Users can then set the desired p-value
and measure of association threshold. The tools then shows a red edge between
all the variables that are correlated in the data but the correlation is not
explained by the current DAG. The thickness of the edge represents the strength
of this correlation, which would help the user in prioritizing edges to add.

The tool also computes the
Shipley's C value at any point.

The edge
weight of the edge represents how high this strength is. 
Based on these shown potential edges, the user can select the edge that
they want to add and decide the direction for it. If the tool detects an
incorrect edge in the graph, it is shown in black. Once the user is happy with
the graph they have they can export it into different formats to use it for
further computation.
